---
output:
  pdf_document: default
  html_document: default
---

# Introduction

## Format
This chapter will provide some background information to put the thesis into context. It will provide information on obesity, in particular its prevalence and implications for cardiovascular disease. Possible mechanisms of disease will be introduced, such as the involvement of platelets. At the end of this chapter, the aims of the thesis will be introduced.

## Background

### Obesity on the rise
Obesity, defined as a body mass index (BMI) of greater than 30kg/m^2^, has tripled worldwide since 1975. It is currently estimated that 40% of adults in the United States and 26% of adults in the United Kingdom have obesity [@Bluher2019]. As of 2016, the number of individuals globally with obesity surpassed 650 million [@TheLancetPublicHealth2018]. The average BMI of the UK adult population is ~27kg/m^2^ [@Wade2018], which is in the 'overweight' category (between 25 and 30 kg/m^2^). Overweight is also more common than 'normal-weight' in many other higher-income countries [@NCD-RisC2016], and prevalence of overweight is increasing in lower-income countries [@Templin2019].

BMI is generally used as a proxy for adiposity as there are high correlations between BMI and objective measures of fat mass measured by dual-energy X-ray absorptiometry (DXA) [@Bell2018], such as total fat mass index and regional fat mass indices. Due to the specialised equipment needed for fat mass measures, BMI is often much more widely available within cohort studies.

### Obesity is a modifiable risk factor for cardiovascular disease
Higher adiposity is a risk factor for certain non-communicable diseases. Such diseases include Type 2 Diabetes mellitus (T2D), musculoskeletal diseases, certain types of cancer and cardiovascular disease [@Garg2014; @Khan2018; @Kortt2002; @Bhaskaran2014]. Collectively, treatment for obesity-related diseases is costly for health services [@Dixon2019a; @Dixon2019]. Cardiovascular disease is one of the leading causes of death. Cardiovascular disease is a general term for diseases affecting the heart or circulatory system and can be subcategorised further into diseases such as coronary heart disease (CHD) and stroke. Obesity is associated with an increased risk of such cardiovascular events [@Khan2018]

### Mendelian randomisation can be used to explore causality
Mendelian randomisation is a technique which uses the random assortment of alleles to estimate the causal association between a modifiable risk factor (such as BMI) and disease outcomes. It can be likened to a natural randomised control trial (RCT) [FIG]. Associations between BMI and disease have been shown through use of prospective observational studies [@Khan2018], but more recently, through use of Mendelian randomisation studies [@Nordestgaard2012; @Dale2017; @Carreras-Torres2017a]. These studies have provided evidence for a causal relationship between adiposity and ischaemic heart disease (IHD) [@Nordestgaard2012], coronary heart disease (CHD), T2D [@Dale2017] and types of cancer including pancreatic cancer [@Carreras-Torres2017a]. Studies have also found that weight loss improves cardiovascular risk factors [@Wing2011; other refs]. Despite the evidence for a causal association between adiposity and cardiovascular disease, mechanisms of obesity-related disease remain unclear.

### Mechanisms of obesity-related cardiovascular disease
Obesity is caused by an increase in fat deposition and generally occurs when energy intake exceeds energy expenditure. There are two types of fat: white adipose tissue (WAT) and brown adipose tissue (BAT). WAT secretes proteins and peptides which have pro-inflammatory effects [@Cercato2019; @Kern2001]. Due to these adipose tissue effects, obesity results in a state of chronic low-grade inflammation [@Esser2014], with an increase in adipose tumour necrosis factor-alpha (TNF-alpha) and interleukin-6 (IL-6) [@Kern2001]. Efforts have been made to understand how adiposity causes cardiovascular disease, with suggestion that inflammation, increased production of reactive oxygen species and endothelial dysfunction contribute to increased cardiovascular risk [@Cercato2019, add other refs].

### Plasma markers of obesity
Studies which have explored the molecular footprint of adiposity have tended to focus on the lipidome, such as cholesterol and triglycerides in lipoprotein subtypes, inlcuding low-density lipoproteins (LDL), very-low-density lipoproteins (VLDL) and high-density lipoproteins (HDL) [@Bell2018a; @Wurtz2014], where a higher BMI is associated with an increase in LDLs and a decrease in large HDLs. This has been expored in a causal framework. 

As the majority of druggable targets are proteins, it is also valuable to explore the effects of BMI on circulating plasma proteins. It is estimated that 25% of proteins in the human proteome circulate in the blood [@Imming2006], including enzymes, protein kinases and transport proteins. Therefore, focusing on effects of adiposity on the circulating proteome may allow for the development of or repurposing of therapeutic interventions to reduce obesity-related disease. Previous studies that have explored the effect of adiposity on the circulating proteome have used hypothesis driven approaches [ref]. Attempts to evaluate the effect of BMI on the general proteome have been performed using observational analyses [refs]. Recent technologies have been developed which allow protein detection in unprecedented scope and detail, for example the SOMAscan by SomaLogic [@Gold2012], which can detect over 3,500 unique proteins. Although identifying alterations of individual proteins may provide therapeutic opportunities, assessing the overall protein profile of adiposity may provide more of a mechanistic insight into how adiposity causes disease.


### Platelets are key cells involved in thrombosis


### Importance of an interdisciplinary approach

### Thesis aims and objectives
The aims of the thesis are

1) 

### Thesis structure
There will be xx results chapters in this thesis.
1)
2)
3)
4)

### Publications from the PhD (in preparation or published)

Redundant role of ASK1-mediated p38MAPK activation in human platelet function. Sledz, K. M., Moore, S. F., Vijayaragavan, V., Mallah, S., **Goudswaard, L. J.**, Williams, C. M., Hunter, R. W. & Hers, I., 1 Apr 2020, In : Cellular Signalling. 68, 11 p., 109528.

Association between blood pressure variability and night time dipping with cardiac structure: cross-sectional study in adolescents

Effects of adiposity on the human plasma proteome: Observational and Mendelian randomization estimates

Estimating the causal association of body mass index and intrinsic platelet properties: evidence for a positive causal relationship with immature platelet count


### Presentations from the PhD

**Platelet Society meeting**, 2018, Cambridge, UK. Title: Do the chemokines MDC and TARC to obesity related platelet hyperactivity and cardiovascular disease? (Oral presentation).

**IEU Scientific Advisory Board meeting**, 2020, Bristol, UK. Title: Effect of adiposity on the human plasma proteome: a Mendelian randomisation study using individual level data (Poster).

**International Genetic Epidemiological Society meeting**, 2020, Virtual conference. Title: Effects of body mass index on the human proteome: a Mendelian randomization study using individual level data (Selected for short talk as one of the top ranked posters).

**International Society of Thrombosis and Haemostasis meeting**, 2020, Virtual conference. Title: Effects of body mass index on the human plasma proteome – potential drivers of obesity-related cardiovascular disease? (Poster). 






### Inline code

If you'd like to put the results of your analysis directly into your discussion, add inline code like this:

> The `cos` of $2 \pi$ is `r cos(2*pi)`. 

Another example would be the direct calculation of the standard deviation:

> The standard deviation of `speed` in `cars` is `r sd(cars$speed)`.

One last neat feature is the use of the `ifelse` conditional statement which can be used to output text depending on the result of an **R** calculation:

> `r ifelse(sd(cars$speed) < 6, "The standard deviation is less than 6.", "The standard deviation is equal to or greater than 6.")`

Note the use of `>` here, which signifies a quotation environment that will be indented.

As you see with `$2 \pi$` above, mathematics can be added by surrounding the mathematical text with dollar signs.  More examples of this are in [Mathematics and Science] if you uncomment the code in [Math].  

### Including plots

You can also embed plots.  For example, here is a way to use the base **R** graphics package to produce a plot using the built-in `pressure` dataset:

```{r pressure, echo=FALSE, cache=TRUE}
plot(pressure)
```

Note that the `echo=FALSE` parameter was added to the code chunk to prevent printing of the **R** code that generated the plot.  There are plenty of other ways to add chunk options.  More information is available at <http://yihui.name/knitr/options/>.  

Another useful chunk option is the setting of `cache=TRUE` as you see here.  If document rendering becomes time consuming due to long computations or plots that are expensive to generate you can use knitr caching to improve performance.  Later in this file, you'll see a way to reference plots created in **R** or external figures.

### Loading and exploring data

Included in this template is a file called `flights.csv`.  This file includes a subset of the larger dataset of information about all flights that departed from Seattle and Portland in 2014.  More information about this dataset and its **R** package is available at <http://github.com/ismayc/pnwflights14>.  This subset includes only Portland flights and only rows that were complete with no missing values.  Merges were also done with the `airports` and `airlines` data sets in the `pnwflights14` package to get more descriptive airport and airline names.

We can load in this data set using the following command:

```{r load_data}
flights <- read.csv("data/flights.csv")
```

The data is now stored in the data frame called `flights` in **R**.  To get a better feel for the variables included in this dataset we can use a variety of functions.  Here we can see the dimensions (rows by columns) and also the names of the columns.

```{r str}
dim(flights)
names(flights)
```

Another good idea is to take a look at the dataset in table form.  With this dataset having more than 50,000 rows, we won't explicitly show the results of the command here.  I recommend you enter the command into the Console **_after_** you have run the **R** chunks above to load the data into **R**.

```{r view_flights, eval=FALSE}
View(flights)
```

While not required, it is highly recommended you use the `dplyr` package to manipulate and summarize your data set as needed.  It uses a syntax that is easy to understand using chaining operations.  Below I've created a few examples of using `dplyr` to get information about the Portland flights in 2014.  You will also see the use of the `ggplot2` package, which produces beautiful, high-quality academic visuals.

We begin by checking to ensure that needed packages are installed and then we load them into our current working environment:

```{r load_pkgs, message=FALSE}
# List of packages required for this analysis
pkg <- c("dplyr", "ggplot2", "knitr", "bookdown", "devtools")
# Check if packages are not installed and assign the
# names of the packages not installed to the variable new.pkg
new.pkg <- pkg[!(pkg %in% installed.packages())]
# If there are any packages in the list that aren't installed,
# install them
if (length(new.pkg))
  install.packages(new.pkg, repos = "http://cran.rstudio.com")
# Load packages (thesisdown will load all of the packages as well)
library(thesisdown)
```

\clearpage

The example we show here does the following:

- Selects only the `carrier_name` and `arr_delay` from the `flights` dataset and then assigns this subset to a new variable called `flights2`. 

- Using `flights2`, we determine the largest arrival delay for each of the carriers.

```{r max_delays}
flights2 <- flights %>% 
  select(carrier_name, arr_delay)
max_delays <- flights2 %>% 
  group_by(carrier_name) %>%
  summarize(max_arr_delay = max(arr_delay, na.rm = TRUE))
```


A useful function in the `knitr` package for making nice tables in _R Markdown_ is called `kable`.  It is much easier to use than manually entering values into a table by copying and pasting values into Excel or LaTeX.  This again goes to show how nice reproducible documents can be! (Note the use of `results="asis"`, which will produce the table instead of the code to create the table.)  The `caption.short` argument is used to include a shorter title to appear in the List of Tables.

```{r maxdelays, results="asis"}
kable(max_delays, 
      col.names = c("Airline", "Max Arrival Delay"),
      caption = "Maximum Delays by Airline",
      caption.short = "Max Delays by Airline",
      longtable = TRUE,
      booktabs = TRUE)
```

The last two options make the table a little easier-to-read.

We can further look into the properties of the largest value here for American Airlines Inc.  To do so, we can isolate the row corresponding to the arrival delay of 1539 minutes for American in our original `flights` dataset.


```{r max_props}
flights %>% filter(arr_delay == 1539, 
                  carrier_name == "American Airlines Inc.") %>%
  select(-c(month, day, carrier, dest_name, hour, 
            minute, carrier_name, arr_delay))
```

We see that the flight occurred on March 3rd and departed a little after 2 PM on its way to Dallas/Fort Worth.  Lastly, we show how we can visualize the arrival delay of all departing flights from Portland on March 3rd against time of departure.

```{r march3plot, fig.height=3, fig.width=6}
flights %>% filter(month == 3, day == 3) %>%
  ggplot(aes(x = dep_time, y = arr_delay)) + geom_point()
```

## Additional resources

- _Markdown_ Cheatsheet - <https://github.com/adam-p/markdown-here/wiki/Markdown-Cheatsheet>

- _R Markdown_ Reference Guide - <https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf>

- Introduction to `dplyr` - <https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html>

- `ggplot2` Documentation - <http://docs.ggplot2.org/current/>
