---
output: 
    thesisdown::thesis_pdf: default
bibliography: bib/library.bib
csl: csl/nature.csl
space_betwee_paragraphs: true
fig_caption: true
always_allow_html: yes
link-citations: true
toc-depth: 2
lot: true
lof: true
indent: true
header-includes: # include other LaTeX packages here
    \usepackage{booktabs}
    \usepackage{longtable}
    \usepackage{siunitx}
    \usepackage[left]{lineno}
    \usepackage{float}
    \floatplacement{figure}{H}
    \linenumbers
    \pagestyle{plain}
    \raggedbottom 
    \usepackage{indentfirst}
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE, message=FALSE}
source("~/OneDrive - University of Bristol/PhD/Main project/Thesis/index/packages.R")
```
# Combining MR and an RCT to determine effects of adiposity on the plasma proteome. {#Comparison-proteome}

## Background

The two previous results chapters (Chapter \@ref(BMI-protein-MR) and Chapter \@ref(BMI-protein-RCT)) aimed to determine the effect of adiposity on circulating proteins by using two different study designs: MR and a RCT which induced weight loss through caloric restriction. Both studies provided evidence that the change in circulating proteins may be contributing to the increased risk of disease such as metabolic disease and cardiovascular disease. This overall shift in proteins provides important information. One of the benefits of having the same proteomics platform in both studies is that there is a large overlap in the number of proteins measured. Therefore, for each protein, it is possible to compare the effect that higher BMI in the general population has, with the effect of short term weight loss.

Aim:
1) Compare BMI-protein effects across MR and RCT study designs (Results from Chapter \@ref(BMI-protein-MR) (Figure \@ref(fig:BMI-protein-graphic2))).

## Methods
In efforts to further elucidate which proteins are altered by adiposity, a comparison between the BMI-protein associations in INTERVAL (Chapter \@ref(BMI-protein-MR)) and the BMI change-protein change associations in the current analyses was also performed using a linear regression. There were 2803 unique proteins (by UniProt ID) which were detected in both INTERVAL and DiRECT. Estimates across the two cohorts were compared using linear regression. 

## Results
### Comparison of BMI effects across DiRECT and INTERVAL
BMI-protein estimates from this study were compared with effect estimates from INTERVAL (Chapter \@ref(BMI-protein-MR)), where 2803 proteins were detected across both studies. Both cohorts, through use of different study designs, provided evidence for an effect of BMI on the plasma proteome. Figure \@ref(fig:DiRECT-INTERVAL) shows the BMI-protein estimates for the proteins which were included in both DiRECT and INTERVAL (R^2^=0.11, P=1.0x10^77^).  There was consistent evidence across both studies that higher BMI leads to higher levels of various proteins including fumarylacetoacetase (FAAA), alcohol dehydrogenase 4 (ADH4), growth hormone receptor (GHR) and carboxypeptidase M (CBPM). Furthermore, higher BMI had a lowering effect on levels of sex hormone binding globulin (SHBG), as well as insulin-like growth factor binding protein 1/2 (IGFBP1/2). Although some of these estimates did not pass multiple testing in INTERVAL, the concordance in estimates in terms of effect size and direction of effect suggests that this is an issue of power.

Nic's comments regarding the figure:
1) overlap and persistent signal
2) heterogeneity and study specific effects
3) strengths of this approach
4) limitations - differenet measures e.g. acute/chronic.

\blandscape
(ref:scaption-DiRECT-INTERVAL) Scatter plot comparing the two stage least squares estimate for the effect of BMI change on protein change in DiRECT with the Mendelian randomisation estimate for the effect of BMI on protein levels in INTERVAL
(ref:caption-DIRECT-INTERVAL) **Scatter plot comparing the two stage least squares estimate for the effect of BMI change on protein change in DiRECT with the Mendelian randomisation estimate for the effect of BMI on protein levels in INTERVAL.** Proteins that are similar across analyses with a large effect size in both studies are coloured in green, whereas those which display similar estimates across studies with small effect size are shown in blue. Proteins do not show similar effects across studies are shown in pink. A linear regression line is indicated in blue.
```{r DiRECT-INTERVAL, echo = FALSE, out.width="90%", fig.cap="(ref:caption-DIRECT-INTERVAL)", fig.scap= "(ref:scaption-DiRECT-INTERVAL)"}
include_graphics(path = "figure/DiRECT/DiRECT_INTERVAL_comparison.pdf", dpi = NA)

```
\elandscape

## Discussion
Comparisons of protein effects across the different study designs (INTERVAL in Chapter \@ref(BMI-protein-MR) and DiRECT in Chapter \@(BMI-protein-RCT)) points towards proteins of interest which may be involved in the development or progression of obesity-associated cardiometabolic disease. Proteins which are only altered in one study design could also provide useful biological information as to how the individual protein may relate to adiposity.

The comparison of results from the previous chapter (Chapter \@ref(BMI-protein-MR)), which used Mendelian randomization and observational associations to assess the effect of average higher BMI across a population on average protein levels, with results from a randomised control trial in the current chapter provides two separate lines of evidence to suggest a causal role of BMI in levels of various proteins. The analysis in INTERVAL suggested that proteins most altered by BMI were enriched for diseases such as cardiovascular disease and metabolic disease. As the estimates for the effect of BMI on proteins were consistent across both DiRECT and INTERVAL, this suggests that weight loss is able to reverse the protein changes observed with higher BMI, thereby potentially reducing disease risk. The enrichment results comparing the 144 proteins which were altered with BMI change to the full 4601 proteins tested confirmed this as there was strongest evidence for enrichment of proteins involved in metabolic disease and renal disease, however the effect on cardiovascular disease did not pass multiple testing adjustment. Pathway enrichment using StringDB also suggested that proteins most altered with BMI change were involved in cytokine-cytokine receptor interactions and complement and coagulation cascades, rather that platelet activation. It could be that the effect on inflammation and coagulation may be more immediate effects, and that changes in platelet activation may either require a more sustained weight loss, or a further reduction in BMI to see effects.

Divergence in estimates across study designs may also provide some insight into the biology due to the nature of each study design. In general, the MR study design provides information on the effect on the proteome that a sustained high BMI has over the life course. In contrast, the estimates in DiRECT are the effect that a relatively short-term weight loss intervention (1 year) has on the proteome. A protein may be altered in the MR study, but not after a weight loss intervention because weight loss may not have been sustained long enough to induce changes at the proteome level. A protein may also not have evidence for a causal relationship in the MR but be altered in DiRECT. This could be because the protein is a signal of starvation or other changes that happen with caloric restriction. 




